The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-27 manuscript.Rmd
move Whole Tale description into Binder use case, closes #38

**Whole&nbsp;Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
For R users, Whole&nbsp;Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.

@@

**\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on **Binder**.
The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
Binder's core internal tool is [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html).
It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files.
In the most powerful case, `repo2docker` builds a given `Dockerfile`.
While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
\pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
\pkg{holepunch} as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
The `Dockerfile` included with the project can also be used to launch a RStudio server locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
The local image usage reduces the number of seperately managed environments and thereby reduces work and increases portability and reproducibility.

>>>>

@@

The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
Binder's core internal tool is `repo2docker`.
It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{https://repo2docker.readthedocs.io/en/latest/config_files.html}.}.
In the most powerful case, `repo2docker` builds a given `Dockerfile`.
While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
This is partly because `repo2docker` does not support arbitrary base images due to the complex auto-generation of the `Dockerfile` instructions.
Two approaches make using Binder easier. for R users.
First, **\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on **Binder**.
For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
\pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
\pkg{holepunch} as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
The `Dockerfile` included with the project can also be used to launch a RStudio server locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
The local image usage reduces the number of seperately managed environments and thereby reduces work and increases portability and reproducibility.
Second, the **Whole&nbsp;Tale** project (https://wholetale.org) combines the strengths of the Rocker Project's curated Docker images with `repo2docker`.
Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
For R users, Whole&nbsp;Tale extends the Jupyter Project's `repo2docker` component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
